---
title: "Predicted vs. Vegas"
author: "Shane Hanley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Introduction

The main goal of my project is to simulate NFL regular season games and use that to generate predictions.  
These probabilities can be converted to predicted moneyline and spreads for the games.  
Just for clarification, the moneyline refers to picking who will win the game with 
the associated odds.  The spread is predicting by much a team needs to at least win
or lose by in a certain game.  The use of these predicted moneylines and spreads
can be compared to the current vegas odds to determine if there is an edge.  
Moreover, if your predicted spread or moneyline is different than vegas that may
indicate that game being worth a wager.  Of course this is only one piece of
determining which games to wager on but can be useful.  



This first part of the code is loading all the necessary packages.
```{r}
#install.packages("tidyverse", type = "binary")
#install.packages("ggrepel", type = "binary")
#install.packages("nflreadr", type = "binary")
#install.packages("nflplotR", type = "binary")
#install.packages("nflseedR")
#install.packages("shiny")
#install.packages("ggplot2")
library(tidyverse)
library(ggrepel)
library(nflreadr)
library(nflplotR)
library(xgboost)
library(nflseedR)
library(shiny)
library(ggplot2)
```



This portion of the code loads the dataset that contains the vegas moneylines and spreads 
```{r}
odds <- nflreadr::load_schedules()
head(odds)
```


We are now filtering on the current season 2022 and renaming the result variable
to difference.

```{r}
#this filter only includes the 2022 season for spreads and moneylines
#2022 season
season <- odds %>%
  filter(season == '2022') %>%
  rename(difference = "result")
head(season)
```


We are using NFLseedR to simulate the outcomes of the 2022 regular season games.
We are simulating each game 10 different times.
```{r}


sim <- nflseedR::simulate_nfl(
  nfl_season = 2022,
  fresh_season = TRUE,
  simulations = 10,
  sims_per_round = 1
)
```



We need to convert the sim list into a dataframe for later use.
We are also renaming some variables.
```{r}
#convert list into a dataframe and rename some variables

my_df <- as.data.frame(sim$game_summary)
my_df <- my_df %>% rename(predicted_spread_home= "result", away_probability = "away_percentage", home_probability="home_percentage")

head(my_df)
```


These if else statements are converting the probabilities from the simulation
dataset into moneyline odds for both the away and home teams.  
```{r}
#code to calculate the ML for home and away teams based on simulation probabilities

my_df$away_predicted_ML = ifelse(my_df$away_probability > 0.5, 
                                    -(my_df$away_probability/(1 -my_df$away_probability))*100,
                                    ((1-my_df$away_probability)/my_df$away_probability)*100)


my_df$home_predicted_ML = ifelse(my_df$home_probability > 0.5, 
                                    -(my_df$home_probability/(1 -my_df$home_probability))*100,
                                    ((1-my_df$home_probability)/my_df$home_probability)*100)
head(my_df)
```


This portion of the code is merging the simulation and vegas spreads and moneylines dataframes.
```{r}
#join ML and spread dataframe with simulation dataframe
join_season_my_df <- merge(season,my_df,by=c("week","game_type","away_team","home_team"))
join_season_my_df
```



To simplify the joined dataframes, we are only including certain variables
for the demonstration.  We are also filtering on only regular season games.
```{r}
#selecting specific variables and filtering dataframe on regular season games

join_season_my_df <- join_season_my_df %>% select(week,away_team,home_team,game_type, game_id, away_score,home_score,away_moneyline,home_moneyline,spread_line,
predicted_spread_home,away_predicted_ML,home_predicted_ML) %>% filter(game_type =="REG")

head(join_season_my_df)
```



To make it easier for creating the shiny app, we are converting dataframe into a long
format using pivot_longer.  We are separating the joined dataframe into three dataframes.
The three dataframes include one for away moneyline, home moneyline, and spread
respectively.  This code below is for the away moneyline.  
```{r}
 #making an away ML specific dataframe and converting to long format 

away_moneyline <- join_season_my_df %>% select(week,away_team,home_team,game_type, game_id, away_score,home_score,away_moneyline,away_predicted_ML)
 
   away_ML <- away_moneyline %>%
  pivot_longer(cols = c("away_moneyline","away_predicted_ML"), names_to = "Source", values_to = "Moneyline_Spread")
  
   head(away_ML)
```



This is the same code as above but this time for the home moneyline.  
```{r}
 #making a home team ML specific dataframe and converting to long format 

home_moneyline <- join_season_my_df %>% select(week,away_team,home_team,game_type, game_id, away_score,home_score,home_moneyline,home_predicted_ML)  
  home_moneyline
  
   home_ML <- home_moneyline %>%
  pivot_longer(cols = c("home_moneyline","home_predicted_ML"), names_to = "Source", values_to = "Moneyline_Spread")
  
   head(home_ML)
```


This piece of code is for converting the spread to long format.  
```{r}
#making a spread specific dataframe and converting to long format   

spread <- join_season_my_df %>% select(week,away_team,home_team,game_type, game_id, away_score,home_score,,spread_line,predicted_spread_home) 
    
    spread_df <-  spread %>%
  pivot_longer(cols = c("spread_line","predicted_spread_home"), names_to = "Source", values_to = "Moneyline_Spread") 
    
    head(spread_df)
```


The next three blocks of code are renaming the moneyline and spread variables to
vegas and predicted.  
```{r}
#rename the Source variable choices 

away_ML <- away_ML %>%
    mutate(Source = recode(Source,away_moneyline = 'Vegas', away_predicted_ML = 'Predicted'))
    
    head(away_ML)
```


```{r}
 #rename the Source variable choices  

home_ML <- home_ML %>%
    mutate(Source = recode(Source,home_moneyline = 'Vegas', home_predicted_ML = 'Predicted'))
    
    head(home_ML)
```


```{r}
 #rename the Source variable choices   

spread_df <- spread_df %>%
    mutate(Source = recode(Source,spread_line = 'Vegas', predicted_spread_home = 'Predicted'))
   
  head(spread_df)
```



This is the first shiny application.  This app creates a bar chart comparing the 
predicted away moneyline to the vegas away moneyline.  There is an object to filter the chart
on a specific game.  
```{r}
#Away ML shiny app

ui <- fluidPage(
  
  # Sidebar layout with a input and output definitions 
  sidebarLayout(
    
    # Inputs
    sidebarPanel(
      
      # Select variable for x-axis
      selectInput(inputId = "game_id", 
                  label = "Game",
                  choices = c(as.character(unique(away_ML$game_id)), "all"),
                  selected = "game_id", multiple = FALSE),
      
       
      
      
      
    
      
    ),  
    
    # Outputs
  mainPanel(
      plotOutput('bar'),
      br(), hr(), br(),
      
    )
  )
)
  


# Define server function required to create the bar chart
server <- function(input, output) {
  
  observe(print(input$game_id))
  
  # Reactive data
  # depends on input$game_id, so wrap in reactive()
  dat <- reactive({
    if (input$game_id != "all")
    {
    away_ML %>% filter(game_id %in% input$game_id)
    } else {
      away_ML
    }
  })
  
  # Create a bar chart object the plot Output function is expecting
  output$bar <- renderPlot({

    ggplot(data = dat(), aes(x = Source, y = Moneyline_Spread)) +
      geom_bar(color = 'black', fill = 'purple', stat='identity') +
      labs(title = paste("Bar Chart of", input$game_id)) +
      geom_text(aes(label = Moneyline_Spread), vjust = -0.2)
    
  
    
    
    
    
  })
  
  
    
  }


# Create a Shiny app object
shinyApp(ui = ui, server = server)
```


This is the second shiny application.  This app creates a bar chart comparing the 
predicted home moneyline to the vegas home moneyline.  There is an object to filter the chart
on a specific game.  
```{r}
#Home ML shiny app

ui <- fluidPage(
  
  # Sidebar layout with a input and output definitions 
  sidebarLayout(
    
    # Inputs
    sidebarPanel(
      
      # Select variable for x-axis
      selectInput(inputId = "game_id", 
                  label = "Game",
                  choices = c(as.character(unique(home_ML$game_id)), "all"),
                  selected = "game_id", multiple = FALSE),
      
       
      
      
      
    
      
    ),  
    
    # Outputs
  mainPanel(
      plotOutput('bar'),
      br(), hr(), br(),
      
    )
  )
)
  


# Define server function required to create the bar chart
server <- function(input, output) {
  
  observe(print(input$game_id))
  
  # Reactive data
  # depends on input$game_id, so wrap in reactive()
  dat <- reactive({
    if (input$game_id != "all")
    {
    home_ML %>% filter(game_id %in% input$game_id)
    } else {
      home_ML
    }
  })
  
  # Create a bar chart object the plo tOutput function is expecting
  output$bar <- renderPlot({

    ggplot(data = dat(), aes(x = Source, y = Moneyline_Spread)) +
      geom_bar(color = 'black', fill = 'purple', stat='identity') +
      labs(title = paste("Bar Chart of", input$game_id)) +
      geom_text(aes(label = Moneyline_Spread), vjust = -0.2)
    
  
    
    
    
    
  })
  
  
    
  }


# Create a Shiny app object
shinyApp(ui = ui, server = server)
```


This is the third shiny application.  This app creates a bar chart comparing the 
predicted spread to the vegas spread.  There is an object to filter the chart
on a specific game.  The predicted spread is for the home team.  The negative predicted
spreads mean the home team is losing by that amount.  The negative vegas spreads would
need to go a step further to align with the moneyline odds.  Comparing the vegas
spread with the vegas moneylines will tell you which team was favored in the spread.  
```{r}
#Spread shiny app

ui <- fluidPage(
  
  # Sidebar layout with a input and output definitions 
  sidebarLayout(
    
    # Inputs
    sidebarPanel(
      
      # Select variable for x-axis
      selectInput(inputId = "game_id", 
                  label = "Game",
                  choices = c(as.character(unique(spread_df$game_id)), "all"),
                  selected = "game_id", multiple = FALSE),
      
       
      
    
    
      
    ),  
    
    # Outputs
  mainPanel(
      plotOutput('bar'),
      br(), hr(), br(),
      
    )
  )
)
  


# Define server function required to create the bar chart
server <- function(input, output) {
  
  observe(print(input$game_id))
  
  # Reactive data
  # depends on input$game_id, so wrap in reactive()
  dat <- reactive({
    if (input$game_id != "all")
    {
    spread_df %>% filter(game_id %in% input$game_id)
    } else {
      spread_df
    }
  })
  
  # Create a bar chart object the plot Output function is expecting
  output$bar <- renderPlot({

    ggplot(data = dat(), aes(x = Source, y = Moneyline_Spread)) +
      geom_bar(color = 'black', fill = 'purple', stat='identity') +
      labs(title = paste("Bar Chart of", input$game_id)) +
      geom_text(aes(label = Moneyline_Spread), vjust = -0.2)
    
  
    
    
    
    
  })
  
  
    
  }


# Create a Shiny app object
shinyApp(ui = ui, server = server)
```

